---
title: "Maternal Mortality Matters"
subtitle: "A Global Insight into Lifetime Risk"
author: "Sophie Ahern"
format:
  html:
    theme: minty
    backgroundcolor: "#e6f4f1"
    toc: true
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(plotly)
library(gridExtra)

# data
Unicef_Metadata_continents_ <- read_csv("Unicef Metadata (continents).csv")
Unicef_Indicator_2 <- read_csv("Unicef Indicator 2.csv")

# transformations
Unicef_data_join_2020 <- Unicef_Metadata_continents_ %>%
  filter(year == 2020)

total_data_join <- full_join(Unicef_data_join_2020, Unicef_Indicator_2)
total_data_join <- full_join(Unicef_data_join_2020, Unicef_Indicator_2, by = c("country" = "country"))
total_data_join <- Unicef_data_join_2020 %>%
  full_join(Unicef_Indicator_2, by = c("country", "year" = "time_period"))
```

Within the complex landscape of global health, the tale of **maternal mortality** unfolds as a poignant narrative. Embark on a visual voyage unveiling a compelling examination of the global **lifetime risk of maternal death**. These visual representations, meticulously crafted using *UNICEF's data*, seek to elevate awareness and ignite initiatives as we delve into the vital domain of maternal welfare.

## Maternal Mortality Map - Navigating Life's Risks

```{r}
#| fig-cap: "This intricate map offers an overview of worldwide maternal mortality rates in 2020."

map_world <- map_data("world")


map_data_join <- full_join(total_data_join, map_world, by = c("country" = "region"))

country_map <- ggplot(map_data_join) +
  aes(x = long, y = lat, group = group, fill = Probability) + 
  geom_polygon() +
  scale_fill_gradient(low = "orange", high = "red", na.value = "grey80") +
   labs(title = "2020 Global Maternal Mortality Probabilities for 15-Year-Old Girls by Country",
       subtitle = "countries shaded in grey have no data due to discrepancies in their names",
        fill = "Maternal Mortality 
        Probabilities") +
  theme(text = element_text(family = "serif"),
        plot.title = element_text(size = 14, face = "bold"),
        plot.background = element_rect(fill = "white", color = "white"),
        panel.background = element_rect(fill = "white", color = "white"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
print(country_map)
```

Each country’s colour reflects its maternal mortality rate, creating striking distinctions. **Vibrant colours** highlight regions with **elevated risks**, while **serene tones** signify areas with **enhanced maternal outcomes**.

## Continental Contrasts - Maternal Mortality Rates

```{r}
#| fig-cap: "This bar chart depicts maternal mortality rates by continent."

map_data_join_continent <- full_join(total_data_join, map_world, by = c("country" = "region"))
map_data_join_continent_mean <- map_data_join_continent %>%
  group_by(continent) %>%
  summarise(mean_indicator = mean(Probability, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(continent))

barchart <- ggplot(map_data_join_continent_mean, aes(x = continent, y = mean_indicator, fill = continent)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Asia" = "lightblue", "Europe" = "yellow", "Africa" = "lightgreen", "Oceania" = "lightpink", "Americas" = "orange")) +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(size = 14, face = "bold"), axis.text.x = element_text(angle = 45), text = element_text(family = "serif")) +
  aes(reorder(continent, mean_indicator)) +
  labs(x = "Continent", y = "Maternal Mortality Probabilities") +
  labs(title = "2020 Average Global Maternal Mortality Probabilities for 15-Year-Old Girls by Continent") +
  guides(colour = "none")

ggplotly(barchart)
```

This bar chart reveals pronounced disparities among regions. Notably, Africa stands out with considerably higher maternal mortality rates compared to other continents, underscoring the **vast gap** between **developed and developing nations**. The visual presentation emphasizes the urgent need for targeted interventions and resources to address maternal health challenges, particularly in African countries.

## Life and Death - Exploring Maternal Health and Longevity

```{r}
#| fig-cap: "This scatterplot depicts the correlation between a country's average life expectancy and its maternal mortality probabilities."

scatter_plot_1 <- total_data_join %>%
  filter(!is.na(continent)) %>%
  ggplot(aes(x = `Life expectancy`, y = Probability, color = continent, size = `Population, total`)) +
  geom_point(alpha = 1) +
  scale_color_manual(values = c("lightblue", "yellow", "lightgreen", "lightpink", "orange")) +
  labs(title = "Relationship between Maternal Mortality Probabilities and Life Expectancy", y = "Maternal Mortality Probabilities", x = "Life Expectancy") +
  theme(text = element_text(family = "serif"), title = element_text(size = 12, face = "bold")) +
  guides(size = "none")

scatter_plotly_1 <- ggplotly(scatter_plot_1)
ggplotly(scatter_plot_1)
```

Within this scatterplot, a **clear pattern** unveils itself, countries with **higher maternal mortality rates** tend to have **lower average life expectancies**, while those with **lower maternal mortality rates** exhibit **higher average life expectancies**. This underscores the crucial role of access to quality maternal care in influencing overall well-being and longevity. Conversely, regions facing higher maternal mortality rates often encounter challenges in healthcare access, education, and socio-economic development. Addressing maternal health emerges as a key component of broader public health efforts, with the potential to improve both maternal well-being and life expectancy, fostering healthier communities.

## Life's Price Tag - Maternal Mortality and GDP vs GNI

```{r}

scatter_plot_2 <- total_data_join %>%
  filter(!is.na(continent)) %>%
  ggplot(aes(x = `GDP per capita`, y = Probability, color = continent, size = `Population, total`)) +
  geom_point(alpha = 1) +
  scale_color_manual(values = c("lightblue", "yellow", "lightgreen", "lightpink", "orange")) +
  labs(title = "Maternal Mortality vs GDP", y = "Maternal Mortality Probability") +  theme(text = element_text(family = "serif"), title = element_text(size = 12, face = "bold")) +
  guides(size = "none")


scatter_plot_3 <- total_data_join %>%
  filter(!is.na(continent)) %>%
  ggplot(aes(x = `GNI`, y = `Probability`, color = continent, size = `Population, total`)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("lightblue", "yellow", "lightgreen", "lightpink", "orange")) +
  labs(title = "Maternal Mortality vs GNI", y = "Maternal Mortality Probabilities") +
  theme(text = element_text(family = "serif"), title = element_text(size = 12, face = "bold")) +
  guides(size = "none") +
  scale_x_continuous(
    limits = c(0, 26000000000000),
    breaks = c(13000000000000))

options(scipen = 999)

grid.arrange(scatter_plot_2, scatter_plot_3, ncol = 2)
```

The examination of maternal mortality rates in the above charts, alongside both Gross National Income (GNI) and Gross Domestic Product (GDP), reveals a **consistent trend** - **wealthier nations** tend to exhibit **lower maternal mortality rates**, whereas **less economically developed nations** often face **higher rates**. This correlation emphasizes the **relationship** between **economic prosperity** and **maternal health outcomes**. 

1. GNI represents the total income earned by a country’s residents globally 
2. GDP focuses solely on domestic economic activity

The inclusion of both GNI and GDP subtly emphasizes how a more encompassing measure, GNI, reflects the global extent of a nation’s wealth, while GDP centres on internal economic production. Despite this distinction, both metrics underscore the crucial connection between economic well-being and maternal health.

## Conclusion and Recommendations

The analysis of maternal mortality rates has underscored a poignant reality, a **stark correlation** exists between **economic disparities**, **life expectancy**, and **maternal health outcomes**. The evidence overwhelmingly suggests that **poorer countries** tend to grapple with **higher maternal mortality probabilities**, amplifying the urgency to address the root causes of these disparities.

In conclusion, addressing maternal mortality rates requires a multifaceted approach that addresses socio-economic, educational, and healthcare disparities. By implementing targeted interventions, such as, 

- prioritizing healthcare infrastructure
- enhancing education initiatives

we can strive towards a future where every woman, regardless of her socio-economic background, has the opportunity to experience a **safe** and **healthy childbirth**.


